---
title: "Usage of Annotation Resources with the CompoundDb Package"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Usage of Annotation Resources with the CompoundDb Package}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{CompoundDb}
    %\VignetteDepends{CompoundDb,RSQLite,Spectra,BiocStyle}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Authors**: `r packageDescription("CompoundDb")[["Author"]] `<br />
**Last modified:** `r file.info("CompoundDb-usage.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r, echo = FALSE, message = FALSE}
library(CompoundDb)
library(BiocStyle)
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```


# Introduction

The `CompoundDb` package provides the functionality to create *compound*
databases from a variety of sources and to use such annotation databases
(`CompDb`). A detailed description on the creation of annotation resources is
given in the *Creating CompoundDb annotation resources* vignette. This vignette
focuses on how annotations can be search for and retrieved.


# Installation

The package (including dependencies) can be installed with the code below:

```{r, eval = FALSE}
install.packages(c("BiocManager", "remotes"))
devtools::install_github("stanstrup/commonMZ")
BiocManager::install("EuracBiomedicalResearch/CompoundDb")
```


# General usage

In this vignette we use a small `CompDb` database containing annotations for a
small number of metabolites build using
[MassBank](https://massbank.eu/MassBank/) release *2020.09*. The respective
`CompDb` database which is loaded below contains in addition to general compound
annotations also MS/MS spectra for these compounds.

```{r load}
library(CompoundDb)
cdb <- CompDb(system.file("sql/CompDb.MassBank.sql", package = "CompoundDb"))
cdb
```

General information about the database can be accessed with the `metadata`
function.

```{r}
metadata(cdb)
```


## Querying compound annotations

The `CompoundDb` package is designed to provide annotation resources for small
molecules, such as metabolites, that are characterized by an exact mass and
additional information such as their IUPAC International Chemical Identifier
[InChI](https://en.wikipedia.org/wiki/International_Chemical_Identifier) or
their chemical formula. The available annotations (*variables*) for compounds
can differ between databases. The `compoundVariables` function can be used to
retrieve a list of all available annotations for compounds.

```{r}
compoundVariables(cdb)
```

The actual compound annotations can then be extracted with the `compounds`
function which returns by default all columns listed by
`compoundVariables`. We can also define specific columns we want to extract with
the `columns` parameter.

```{r}
head(compounds(cdb, columns = c("name", "formula", "exactmass")))
```

As a technical detail, `CompDb` databases follow a very simple database layout
with only few constraints to allow data import and representation for a variety
of sources (e.g. MassBank, HMDB, MoNa, ChEBI). For the present database, which
is based on MassBank, the mapping between entries in the *ms_compound* database
table and MS/MS spectra is for example 1:1 and the *ms_compound* table contains
thus highly redundant information. Thus, if we would include the column
`"compound_id"` in the query we would end up with redundant values instead:

```{r}
head(compounds(cdb, columns = c("compound_id", "name", "formula")))
```

By default, `compounds` extracts the data for **all** compounds stored in the
database. The function supports however also *filters* to get values for
specific entries only. These can be defined as *filter expressions* which are
similar to the way how e.g. a `data.frame` would be subsetted in R. In the
example below we

```{r}
compounds(cdb, columns = c("compound_id", "name", "formula"),
          filter = ~ name == "Mellein")
```

Note that a filter expression always has to start with `~` followed by the
*variable* on which the data should be subsetted and the condition to select the
entries of interest. The list of available filters for the `CompDb` can be
retrieved with the `supportedFilter` function (more filters will be implemented
in future).

```{r}
supportedFilters(cdb)
```

Also, filters can be combined to create more specific filters in the same manner
this would be done in R, i.e. using `&` for *and*, `|` for *or* and `!` for
*not*. To illustrate this we extract below all compound entries from the table
for compounds with the name *Mellein* and that have a `"compound_id"` which is
either 1 or 5.

```{r}
compounds(cdb, columns = c("compound_id", "name", "formula"),
          filter = ~ name == "Mellein" & compound_id %in% c(1, 5))
```

Similarly, we can define a filter expression to retrieve compounds with an exact
mass between 310 and 320.

```{r}
compounds(cdb, columns = c("name", "exactmass"),
          filter = ~ exactmass > 310 & exactmass < 320)
```

In addition to *filter expressions*, we can also define and combine filters
using the actual filter classes. This provides additional conditions that would
not be possible with regular filter expressions. Below we fetch for examples
only compounds from the database that contain a *H14* in their formula. To this
end we use a `FormulaFilter` with the condition `"contains"`. Note that all
filters that base on character matching (i.e. `FormulaFilter`, `InchiFilter`,
`InchikeyFilter`, `NameFilter`) support as conditions also `"contains"`,
`"startsWith"` and `"endsWith"` in addition to `"="` and `"!="`.

```{r}
compounds(cdb, columns = c("name", "formula", "exactmass"),
          filter = FormulaFilter("H14", "contains"))
```

It is also possible to combine filters if they are defined that way, even if it
is a little less straight forward than with the filter expressions. Below we
combine the `FormulaFilter` with the `ExactmassFilter` to retrieve only
compounds with an `"H14"` in their formula and an exact mass between 310 and
320.

```{r}
filters <- AnnotationFilterList(
    FormulaFilter("H14", "contains"),
    ExactmassFilter(310, ">"),
    ExactmassFilter(320, "<"),
    logicOp = c("&", "&"))
compounds(cdb, columns = c("name", "formula", "exactmass"),
          filter = filters)
```


## Extracting with MS/MS data

`CompDb` database can also store and provide MS/MS spectral data. These can be
interfaced with a `Spectra` object from the `r Biocpkg("Spectra")` Bioconductor
package which can be initialized with the `Spectra` function as shown below.

```{r}
sps <- Spectra(cdb)
sps
```

With the `spectraVariables` function it is possible to list all available
annotations specific to a spectrum from the database.

```{r}
spectraVariables(sps)
```

Individual variables can then be accessed with `$` and the variable name:

```{r}
head(sps$adduct)
```

For more information on how to use `Spectra` objects in your analysis have also
a look at the package
[vignette](https://rformassspectrometry.github.io/Spectra/articles/Spectra.html)
or a [tutorial](https://jorainer.github.io/SpectraTutorials/) on how to perform
MS/MS spectra matching with `Spectra`.

Similar to the `compounds` function, a call to `Spectra` will give access to
**all** spectra in the database. Using the same filtering framework it is
however also possible to *extract* only specific spectra from the
database. Below we are for example accessing only the MS/MS spectra of the
compound *Mellein*.

```{r}
mellein <- Spectra(cdb, filter = ~ name == "Mellein")
mellein
```

Instead of all spectra we extracted now only a subset of `r length(mellein)`
spectra from the database.

As a simple toy example we perform next pairwise spectra comparison between the
5 spectra from *Mellein* with all the MS/MS spectra in the database.

```{r}
library(Spectra)
cormat <- compareSpectra(mellein, sps, ppm = 40)
```


# Session information