---
title: "Creating CompoundDb annotation resources"
package: CompoundDb
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Creating CompoundDb annotation resources}
  %\VignetteEngine{knitr::rmarkdown}
  %%\VignetteKeywords{Mass Spectrometry, Metabolomics, Infrastructure, Bioinformatics}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

**Authors**: Johannes Rainer<br />
**Modified**: `r file.info("create-compounddb.Rmd")$mtime`<br />
**Compiled**: `r date()`

# Introduction

Chemical compound annotation and information can be retrieved from a variety of
sources including [HMDB](http://www.hmdb.ca),
[LipidMaps](http://www.lipidmaps.org) or [ChEBI](http://ebi.ac.uk/chebi). The
`CompoundDb` package provides functionality to extract data relevant for
(chromatographic) peak annotations in metabolomics/lipidomics experiments from
these sources and to store it into a common format (i.e. an `CompDb`
object/database). This vignette describes how such `CompDb` objects can be
created exemplified with package-internal test files that represent data subsets
from some annotation resources.

The R object to represent the compound annotation is the `CompDb` object. Each
object (respectively its database) is supposed to contain and provide
annotations from a single source (e.g. HMDB or LipidMaps) but it is also
possible to create cross-source databases too.

# Create a `CompDb` object for HMDB

The `CompDb` package provides the `compound_tbl_sdf` and the
`compound_tbl_lipidblast` functions to extract relevant compound annotation from
files in SDF (structure-data file) format or in the json files from LipidBlast
(http://mona.fiehnlab.ucdavis.edu/downloads). `CompoundDb` allows to process SDF
files from:

- Human Metabolome Database (HMDB), http://www.hmdb.ca
- Chemical Entities of Biological Interest (ChEBI): http://ebi.ac.uk/chebi
- LIPID MAPS Structure Database (LMSD): http://www.lipidmaps.org
- PubChem: https://pubchem.ncbi.nlm.nih.gov

Below we use the `compound_tbl_sdf` to extract compound annotations from a SDF
file representing a very small subset of the HMDB database.

```{r compound_tbl, message = FALSE, warnings = FALSE}
library(CompoundDb)

## Locate the file
hmdb_file <- system.file("sdf/HMDB_sub.sdf", package = "CompoundDb")
## Extract the data
cmps <- compound_tbl_sdf(hmdb_file)
```

The function returns by default a (`data.frame`-equivalent) `tibble` (from the
*tidyverse*'s `tibble` package).

```{r cmps}
cmps
```

The `tibble` contains columns

- `compound_id`: the resource-specific ID of the compound.
- `compound_name`: the name of the compound, mostly a generic or common name.
- `inchi`: the compound's inchi.
- `formula`: the chemical formula of the compound.
- `mass`: the compounds (monoisotopic) mass.
- `synonyms`: a `list` of aliases/synonyms for the compound.

We can pass this `tibble` along with additional required metadata information to
the `createCompDb` function to create a SQLite-based database with the compound
annotations. The metadata information for such a `CompDb` resource is crucial as
it defines the origin of the annotations and its version. This information
should thus be carefully defined by the user. Below we use the `make_metadata`
helper function to create a `data.frame` in the expected format. The organism
should be provided in the format e.g. `"Hsapiens"` for human or `"Mmusculus"`
for mouse, i.e. capital first letter followed by lower case characters without
whitespaces.

```{r metadata, message = FALSE}
metad <- make_metadata(source = "HMDB", url = "http://www.hmdb.ca",
                       source_version = "4.0", source_date = "2017-09",
                       organism = "Hsapiens")
```

We create now a SQLite database for the HMDB subset. With `path` we specify the
path to the directory in which we want to save the database. This defaults to
the current location, but for this example we save the database into a temporary
folder.

```{r createCompDb}
db_file <- createCompDb(cmps, metadata = metad, path = tempdir())
```

To use this database we pass it to the `CompDb` function which returns a
`CompDb` objects acting as an interface to the database.

```{r CompDb}
cmpdb <- CompDb(db_file)
cmpdb
```

We can e.g. use the `compounds` function to extract all compounds from the
database. With parameter `columns` we can choose the database columns to return.

```{r compounds}
compounds(cmpdb, columns = c("compound_name", "formula", "mass"))
```

For use of the created `CompDb` object/database across R sessions or to share
the resouce, we can also create an R-package containing the annotation.  To
enable reproducible research, each `CompDb` package should contain the version
of the originating data source in its file name (which is by default extracted
from the metadata of the resource). Below we create a `CompDb` package from the
generated database file. We have to provide in addition some general information
related to the package such as its version and its maintainer.

```{r createCompoundDbPackage, warning = FALSE}
createCompDbPackage(
    db_file, version = "0.0.1", author = "J Rainer", path = tempdir(),
    maintainer = "Johannes Rainer <johannes.rainer@eurac.edu>")
```

The function creates a folder (in our case in a temporary directory) that can be
build and installed with `R CMD build` and `R CMD INSTALL`.

Special care should also be put on the license of the package that can be passed
with the `license` parameter. The license of the package and how and if the
package can be distributed will depend also on the license of the originating
resource.

# Session information

```{r sessioninfo, echo=FALSE}
sessionInfo()
```

