---
title: "Creating CompoundDb annotation resources"
package: CompoundDb
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Creating CompoundDb annotation resources}
  %\VignetteEngine{knitr::rmarkdown}
  %%\VignetteKeywords{Mass Spectrometry, Metabolomics, Infrastructure, Bioinformatics}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

**Authors**: Johannes Rainer<br />
**Modified**: `r file.info("create-compounddb.Rmd")$mtime`<br />
**Compiled**: `r date()`

# Introduction

Chemical compound annotation and information can be retrieved from a variety of
sources including [HMDB](http://www.hmdb.ca),
[LipidMaps](http://www.lipidmaps.org) or [ChEBI](http://ebi.ac.uk/chebi). The
`CompoundDb` package provides functionality to extract data relevant for
(chromatographic) peak annotations in metabolomics/lipidomics experiments from
these sources and to store it into a common format (i.e. an `CompDb`
object/database). This vignette describes how such `CompDb` objects can be
created exemplified with package-internal test files that represent data 
subsets from some annotation resources.

The R object to represent the compound annotation is the `CompDb` object. Each
object (respectively its database) is supposed to contain and provide
annotations from a single source (e.g. HMDB or LipidMaps) but it is also
possible to create cross-source databases too.


# Create a `CompDb` object for HMDB

The `CompDb` package provides the `compound_tbl_sdf` and the
`compound_tbl_lipidblast` functions to extract relevant compound annotation from
files in SDF (structure-data file) format or in the json files from LipidBlast
(http://mona.fiehnlab.ucdavis.edu/downloads). `CompoundDb` allows to process SDF
files from:

- Human Metabolome Database (HMDB), http://www.hmdb.ca
- Chemical Entities of Biological Interest (ChEBI): http://ebi.ac.uk/chebi
- LIPID MAPS Structure Database (LMSD): http://www.lipidmaps.org
- PubChem: https://pubchem.ncbi.nlm.nih.gov

Below we use the `compound_tbl_sdf` to extract compound annotations from a SDF
file representing a very small subset of the HMDB database.

```{r compound_tbl, message = FALSE, warnings = FALSE}
library(CompoundDb)

## Locate the file
hmdb_file <- system.file("sdf/HMDB_sub.sdf", package = "CompoundDb")
## Extract the data
cmps <- compound_tbl_sdf(hmdb_file)
```

The function returns by default a (`data.frame`-equivalent) `tibble` (from the
*tidyverse*'s `tibble` package).

```{r cmps}
cmps
```

The `tibble` contains columns

- `compound_id`: the resource-specific ID of the compound.
- `compound_name`: the name of the compound, mostly a generic or common name.
- `inchi`: the compound's inchi.
- `formula`: the chemical formula of the compound.
- `mass`: the compounds (monoisotopic) mass.
- `synonyms`: a `list` of aliases/synonyms for the compound.

To create a simple compound database, we could pass this `tibble` along with
additional required metadata information to the `createCompDb` function. In the
present example we want to add however also MS/MS spectrum data to the
database. We thus load below the MS/MS spectra for some of the compounds from
the respective xml files downloaded from HMDB. To this end we pass the path to
the folder in which the files are located to the `msms_spectra_hmdb`
function. The function identifies the xml files containing MS/MS spectra based
on their their file name and loads the respective spectrum data. The folder can
therefore also contain other files, but the xml files from HMDB should not be
renamed or the function will not recognice them. Note also that at present only
MS/MS spectrum xml files from HMDB are supported (one xml file per spectrum).

```{r msms_spectra, message = FALSE}
## Locate the folder with the xml files
xml_path <- system.file("xml", package = "CompoundDb")
spctra <- msms_spectra_hmdb(xml_path)
```

At last we have to create the metadata for the resource. The metadata
information for a `CompDb` resource is crucial as it defines the origin of the
annotations and its version. This information should thus be carefully defined
by the user. Below we use the `make_metadata` helper function to create a
`data.frame` in the expected format. The organism should be provided in the
format e.g. `"Hsapiens"` for human or `"Mmusculus"` for mouse, i.e. capital
first letter followed by lower case characters without whitespaces.

```{r metadata, message = FALSE}
metad <- make_metadata(source = "HMDB", url = "http://www.hmdb.ca",
                       source_version = "4.0", source_date = "2017-09",
                       organism = "Hsapiens")
```

With all the required data ready we create the SQLite database for the HMDB
subset. With `path` we specify the path to the directory in which we want to
save the database. This defaults to the current working directory, but for this
example we save the database into a temporary folder.

```{r createCompDb}
db_file <- createCompDb(cmps, metadata = metad, msms_spectra = spctra,
                        path = tempdir())
```

The variable `db_file` is now the file name of the SQLite database. We can pass
this file name to the `CompDb` function to get the `CompDb` objects acting as
the interface to the database.

```{r CompDb}
cmpdb <- CompDb(db_file)
cmpdb
```

To extract all compounds from the database we can use the `compounds`
function. The parameter `columns` allows to choose the database columns to
return.

```{r compounds}
compounds(cmpdb, columns = c("compound_name", "formula", "mass"))
```

Analogously we can use the `spectra` function to extract spectrum data from the
database. The function returns by default a `Spectrum2List` object, that allows
to store multiple `Spectrum2` objects in a `list`-like structure with additional
annotations available in *metadata columns* (that can be accessed either with
`$` or `mcols`).

```{r spectra}
spectra(cmpdb)
```

To share or archive the such created `CompDb` database, we can also create a
dedicated R package containing the annotation. To enable reproducible research,
each `CompDb` package should contain the version of the originating data source
in its file name (which is by default extracted from the metadata of the
resource). Below we create a `CompDb` package from the generated database
file. Required additional information we have to provide to the function are the
package creator/maintainer and its version.

```{r createCompoundDbPackage, warning = FALSE}
createCompDbPackage(
    db_file, version = "0.0.1", author = "J Rainer", path = tempdir(),
    maintainer = "Johannes Rainer <johannes.rainer@eurac.edu>")
```

The function creates a folder (in our case in a temporary directory) that can be
build and installed with `R CMD build` and `R CMD INSTALL`.

Special care should also be put on the license of the package that can be passed
with the `license` parameter. The license of the package and how and if the
package can be distributed will depend also on the license of the originating
resource.

# Session information

```{r sessioninfo, echo=FALSE}
sessionInfo()
```

